---
title: "Getting Started with R & RStudio for Statistics"
subtitle: "A practical introduction to working with data and making plots in R"
author: "Talissa Floriani"
date: "2026-08-31"

format:
  html:
    toc: true
    toc-title: "In this guide"
    toc-location: right
    toc-depth: 2
    code-copy: true
    code-fold: false
    code-overflow: wrap
    smooth-scroll: true

execute:
  warning: false
  message: false
---

<style>

/* =========================================================
   COLORS
   ========================================================= */

:root {
  --illinois-navy: #13294B;
  --illinois-orange: #FF5F05;

  --pink: #D94F8A;
  --pink-dark: #A93368;
  --pink-light: #FFF1F6;

  --orange-light: #FFF3EC;
  --navy-light: #EFF3F8;

  --cream: #FFFCF8;
  --white: #FFFFFF;
  --text: #25282D;
  --muted: #667085;
  --border: #E4E7EC;
}


/* =========================================================
   PAGE
   ========================================================= */

body {
  background-color: var(--cream);
  color: var(--text);

  font-family:
    "Avenir Next",
    "Segoe UI",
    Helvetica,
    Arial,
    sans-serif;

  font-size: 1.03rem;
  line-height: 1.75;
}

main.content {
  max-width: 880px;
}


/* =========================================================
   TITLE
   ========================================================= */

.quarto-title-block {
  margin-bottom: 2.5rem;
  padding-bottom: 1.5rem;

  border-bottom: 4px solid var(--illinois-orange);
}

.quarto-title h1.title {
  color: var(--illinois-navy);

  font-size: 3rem;
  font-weight: 800;
  line-height: 1.08;

  letter-spacing: -0.04em;
}

.quarto-title .subtitle {
  color: var(--muted);

  font-size: 1.2rem;
  line-height: 1.55;

  max-width: 720px;
}


/* =========================================================
   HEADINGS
   ========================================================= */

h1,
h2,
h3 {
  color: var(--illinois-navy);
}

h1 {
  font-weight: 800;

  letter-spacing: -0.025em;

  margin-top: 3.5rem;
}

h2 {
  font-weight: 750;
  margin-top: 2.3rem;
}

h3 {
  color: var(--pink-dark);
  font-weight: 700;
}


/* =========================================================
   HERO
   ========================================================= */

.hero {
  background:
    linear-gradient(
      135deg,
      #EEF3F9 0%,
      #FFF2F7 55%,
      #FFF5EE 100%
    );

  border: 1px solid var(--border);

  border-radius: 20px;

  padding: 2rem 2.2rem;

  margin:
    1rem 0
    2.8rem 0;
}

.hero h2 {
  color: var(--illinois-navy);

  font-size: 1.8rem;

  margin-top: 0.5rem;
}

.hero-label {
  display: inline-block;

  background: var(--illinois-navy);
  color: white;

  border-radius: 100px;

  padding:
    0.3rem
    0.75rem;

  font-size: 0.72rem;
  font-weight: 750;

  letter-spacing: 0.08em;
}

.pill-row {
  display: flex;
  flex-wrap: wrap;

  gap: 0.5rem;

  margin-top: 1.3rem;
}

.pill {
  background: rgba(255,255,255,0.8);

  color: var(--illinois-navy);

  border: 1px solid var(--border);

  border-radius: 100px;

  padding:
    0.35rem
    0.8rem;

  font-size: 0.83rem;
}


/* =========================================================
   SECTION LABEL
   ========================================================= */

.section-label {
  display: inline-block;

  color: var(--illinois-orange);

  font-size: 0.7rem;
  font-weight: 800;

  letter-spacing: 0.12em;

  margin-top: 2.8rem;
  margin-bottom: -0.3rem;
}

.section-label-pink {
  color: var(--pink);
}


/* =========================================================
   CARDS
   ========================================================= */

.card-grid {
  display: grid;

  grid-template-columns:
    repeat(2, minmax(0, 1fr));

  gap: 1rem;

  margin:
    1.4rem 0
    2.5rem 0;
}

.info-card {
  background: var(--white);

  border: 1px solid var(--border);

  border-radius: 14px;

  padding: 1.3rem 1.4rem;
}

.info-card h3 {
  color: var(--illinois-navy);

  margin-top: 0;
  margin-bottom: 0.35rem;

  font-size: 1.08rem;
}

.info-card p:last-child {
  margin-bottom: 0;
}


/* =========================================================
   R vs RSTUDIO
   ========================================================= */

.compare-grid {
  display: grid;

  grid-template-columns:
    repeat(2, minmax(0, 1fr));

  gap: 1rem;

  margin:
    1.4rem 0
    2rem 0;
}

.compare-card {
  border-radius: 16px;

  padding: 1.5rem;

  border: 1px solid var(--border);
}

.compare-r {
  background: var(--illinois-navy);
  color: white;
}

.compare-r h3 {
  color: white;
  margin-top: 0;
}

.compare-rstudio {
  background: var(--pink-light);
  color: var(--illinois-navy);
}

.compare-rstudio h3 {
  color: var(--pink-dark);
  margin-top: 0;
}


/* =========================================================
   SMALL NOTES
   ========================================================= */

.note-box {
  background: var(--navy-light);

  border-left: 5px solid var(--illinois-navy);

  border-radius:
    0 10px 10px 0;

  padding:
    1rem
    1.25rem;

  margin:
    1.5rem 0;
}

.tip-box {
  background: var(--pink-light);

  border-left: 5px solid var(--pink);

  border-radius:
    0 10px 10px 0;

  padding:
    1rem
    1.25rem;

  margin:
    1.5rem 0;
}

.warning-box {
  background: var(--orange-light);

  border-left: 5px solid var(--illinois-orange);

  border-radius:
    0 10px 10px 0;

  padding:
    1rem
    1.25rem;

  margin:
    1.5rem 0;
}


/* =========================================================
   CODE
   ========================================================= */

div.sourceCode {
  border: 1px solid #E1E5EA;

  border-left:
    4px solid var(--illinois-orange);

  border-radius: 10px;

  overflow: hidden;
}

pre.sourceCode {
  padding: 1rem;
}

code {
  font-family:
    "SFMono-Regular",
    Consolas,
    "Liberation Mono",
    monospace;
}

:not(pre) > code {
  background: #F1F3F6;

  color: var(--pink-dark);

  border-radius: 4px;

  padding:
    0.1rem
    0.3rem;
}


/* =========================================================
   GGPLOT FORMULA
   ========================================================= */

.ggplot-formula {
  display: flex;

  align-items: center;

  justify-content: center;

  flex-wrap: wrap;

  gap: 0.55rem;

  background: var(--white);

  border: 1px solid var(--border);

  border-radius: 14px;

  padding: 1.2rem;

  margin:
    1.7rem 0;
}

.formula-part {
  background: var(--navy-light);

  color: var(--illinois-navy);

  padding:
    0.55rem
    0.8rem;

  border-radius: 8px;

  font-size: 0.77rem;
  font-weight: 800;

  letter-spacing: 0.04em;
}

.formula-plus {
  color: var(--pink);

  font-size: 1.2rem;
  font-weight: 800;
}


/* =========================================================
   TABLES
   ========================================================= */

table {
  width: 100%;

  background: white;

  border-radius: 10px;

  overflow: hidden;

  margin:
    1.5rem 0;
}

thead {
  background: var(--illinois-navy);
  color: white;
}

thead th {
  color: white;
}

tbody tr:nth-child(even) {
  background: #F8F9FB;
}


/* =========================================================
   LINKS
   ========================================================= */

a {
  color: var(--pink-dark);

  text-underline-offset: 3px;
}

a:hover {
  color: var(--illinois-orange);
}


/* =========================================================
   TOC
   ========================================================= */

#TOC {
  font-size: 0.88rem;
}

#TOC .active {
  color: var(--pink) !important;
  font-weight: 700;
}

.sidebar nav[role="doc-toc"] ul > li > a:hover {
  color: var(--illinois-orange);
}


/* =========================================================
   PRACTICE
   ========================================================= */

.practice-box {
  background:
    linear-gradient(
      135deg,
      #FFF1F6,
      #FFF8F2
    );

  border: 1px solid #F0D7E1;

  border-radius: 16px;

  padding: 1.5rem;

  margin:
    1.5rem 0;
}


/* =========================================================
   FINAL BOX
   ========================================================= */

.final-box {
  background: var(--illinois-navy);

  color: white;

  border-radius: 16px;

  padding: 1.7rem;

  margin:
    2rem 0;
}

.final-box h2,
.final-box h3 {
  color: white;
}


/* =========================================================
   MOBILE
   ========================================================= */

@media (max-width: 768px) {

  .quarto-title h1.title {
    font-size: 2.3rem;
  }

  .card-grid,
  .compare-grid {
    grid-template-columns: 1fr;
  }

  .hero {
    padding: 1.4rem;
  }

}

</style>


```{r}
#| label: setup
#| include: false

library(ggplot2)

plants <- data.frame(
  plant = 1:20,

  treatment = rep(
    c("Control", "Fertilizer"),
    each = 10
  ),

  height = c(
    12.1, 11.5, 13.2, 12.8, 11.9,
    12.6, 13.4, 12.2, 11.8, 13.0,
    15.2, 14.8, 16.1, 15.5, 14.9,
    15.7, 16.4, 14.6, 15.1, 16.0
  ),

  biomass = c(
    4.1, 3.8, 4.4, 4.2, 4.0,
    4.3, 4.5, 4.0, 3.9, 4.3,
    5.2, 5.0, 5.6, 5.3, 5.1,
    5.5, 5.8, 4.9, 5.2, 5.6
  )
)
```


::: {.hero}

<div class="hero-label"> Documented instructions developed for students at the CPSC 440 - UIUC </div>

## New to R? Start here.

This guide covers the R basics I think are most useful when you're starting a statistics class: how RStudio is organized, how data are stored, a few functions you'll use all the time, and how to make your first plots with `ggplot2`. 

You don't need any previous coding experience.

<div class="pill-row">

<span class="pill">Beginner level</span>

<span class="pill">Statistics focused</span>

<span class="pill">ggplot2 included</span>

</div>

:::


## What we'll cover

<div class="card-grid">

<div class="info-card">

### RStudio basics

Where to write code, where R runs it, and where to find your files, plots, and objects.

</div>

<div class="info-card">

### Working with data

Objects, vectors, data frames, missing values, and a few basic ways to inspect a dataset.

</div>

<div class="info-card">

### Descriptive statistics

Means, medians, standard deviations, ranges, and summaries.

</div>

<div class="info-card">

### Data visualization

The basic structure of `ggplot2`, plus scatterplots, histograms, and boxplots.

</div>

</div>


<div class="section-label">STARTING WITH THE BASICS</div>

# R and RStudio are different things

This distinction is useful from the beginning.

<div class="compare-grid">

<div class="compare-card compare-r">

### R

**R is the language doing the work.**

It performs the calculations, fits the models, manipulates the data, and creates the figures.

For example:

```r
mean(height)
```

</div>

<div class="compare-card compare-rstudio">

### RStudio

**RStudio is the interface we use to work with R.**

It gives us a place to write scripts, run code, view plots, inspect datasets, and organize files.

</div>

</div>

A simple way to remember it is:

> **R is the engine; RStudio is the workspace around it.**


# Getting familiar with RStudio

When you first open RStudio, there are usually four main areas.

## Source

The **Source** pane is where you write and save scripts.

For example, an R script might contain:

```r
height <- c(10, 12, 15, 13, 11)

mean(height)
sd(height)
```

This is where most of your analysis should live.

## Console

The **Console** is where R executes commands.

Try:

```{r}
5 * 10
```

You can type commands directly into the Console, but they are not automatically saved as part of your analysis.

::: {.warning-box}

I use the Console for quick checks, but I would not recommend doing an entire homework assignment or analysis there. Put code you want to keep in a script.

:::

## Environment

The **Environment** shows the objects currently stored in your R session.

For example:

```{r}
my_number <- 42
```

After running that line, `my_number` should appear in the Environment.

## Files, Plots, Packages, and Help

The lower-right panel has a few tabs you will use often.

- **Files** — navigate through your project files
- **Plots** — view figures created in R
- **Packages** — see installed packages
- **Help** — open R documentation


# Start with an RStudio Project

For class work, I recommend creating an RStudio Project for each course or analysis.

Go to:

> **File → New Project → New Directory → New Project**

You might call it:

```text
statistics-class
```

A simple project could look like:

```text
statistics-class/
│
├── data/
│   └── experiment.csv
│
├── figures/
│
├── scripts/
│   └── analysis.R
│
└── statistics-class.Rproj
```

::: {.tip-box}

Using a Project helps keep your code and data together and makes it much easier to come back to an analysis later.

It also saves you from file paths like:

```text
/Users/name/Desktop/statistics/final/final2/REAL_final/data.csv
```

:::


<div class="section-label">R BASICS</div>

# Using R as a calculator

At the simplest level, R can do regular arithmetic.

```{r}
2 + 2
```

```{r}
10 - 5
```

```{r}
10 * 5
```

```{r}
10 / 5
```

Exponents use `^`:

```{r}
2^3
```

And parentheses work the same way they do in regular mathematics:

```{r}
(10 + 5) * 2
```

Compare that with:

```{r}
10 + 5 * 2
```

R follows the usual order of operations.


# Creating objects

An **object** stores something so that we can use it again later.

For example:

```{r}
x <- 10
```

The `<-` symbol is the **assignment operator**.

I usually read:

```r
x <- 10
```

as:

> Put `10` into an object called `x`.

Now we can call that object:

```{r}
x
```

Create another one:

```{r}
y <- 5
```

and use them together:

```{r}
x + y
```

```{r}
x * y
```


## Objects can contain different types of information

Numbers:

```{r}
age <- 25
```

Text:

```{r}
name <- "Maria"
```

Logical values:

```{r}
student <- TRUE
```

`TRUE` and `FALSE` are logical values in R. They become especially useful later when we start filtering data.


# A note on object names

You can technically call objects almost anything, but meaningful names make code much easier to read.

This:

```r
plant_height <- 15
```

is more informative than:

```r
x <- 15
```

I usually prefer lowercase names separated with underscores:

```text
plant_height
plant_biomass
treatment_group
mean_height
```

::: {.warning-box}

**R is case-sensitive.**

These are three different names:

```text
height
Height
HEIGHT
```

If you create `height` and later type `Height`, R will not treat them as the same object.

:::


# Vectors

Statistics almost always involves more than one observation.

A **vector** stores several values of the same general type.

Suppose we measured the height of five plants:

```{r}
height <- c(
  12.1,
  11.5,
  13.2,
  12.8,
  11.9
)
```

The `c()` function means **combine**.

Run:

```{r}
height
```

and R prints all five values.


<div class="section-label">DESCRIPTIVE STATISTICS</div>

# Summarizing a vector

Once the measurements are stored in R, we can calculate basic descriptive statistics directly.

## Mean

```{r}
mean(height)
```

## Median

```{r}
median(height)
```

## Standard deviation

```{r}
sd(height)
```

## Variance

```{r}
var(height)
```

## Minimum and maximum

```{r}
min(height)
max(height)
```

## Range

```{r}
range(height)
```

## Number of observations

```{r}
length(height)
```

## A quick summary

```{r}
summary(height)
```

`summary()` is especially useful for getting a quick look at a numeric variable.

::: {.note-box}

Before jumping into a statistical test, spend a little time looking at the data.

At minimum, I usually want to know:

- the center of the distribution;
- how variable the observations are;
- the range;
- whether values are missing;
- whether anything looks unusual.

:::


# Functions and arguments

Commands such as:

```text
mean()
sd()
summary()
```

are **functions**.

A function takes an input and does something with it.

For example:

```{r}
mean(height)
```

applies `mean()` to `height`.

Functions can also have additional arguments.

```{r}
round(
  mean(height),
  digits = 2
)
```

Here, `digits = 2` tells `round()` to keep two decimal places.


# Missing values

Missing values are common in real datasets.

R represents a missing value as:

```text
NA
```

For example:

```{r}
height_missing <- c(
  12.1,
  11.5,
  NA,
  12.8,
  11.9
)
```

Now try:

```{r}
mean(height_missing)
```

R returns `NA`.

To calculate the mean using the available observations:

```{r}
mean(
  height_missing,
  na.rm = TRUE
)
```

`na.rm = TRUE` tells the function to remove missing values before doing the calculation.


<div class="section-label">WORKING WITH DATA</div>

# Data frames

Most statistical datasets contain several variables.

In R, one common way to store this type of dataset is a **data frame**.

A useful way to think about it is:

> **Rows are observations. Columns are variables.**

Let's create a small plant experiment.

```{r}
plants <- data.frame(

  plant = 1:20,

  treatment = rep(
    c("Control", "Fertilizer"),
    each = 10
  ),

  height = c(
    12.1, 11.5, 13.2, 12.8, 11.9,
    12.6, 13.4, 12.2, 11.8, 13.0,
    15.2, 14.8, 16.1, 15.5, 14.9,
    15.7, 16.4, 14.6, 15.1, 16.0
  ),

  biomass = c(
    4.1, 3.8, 4.4, 4.2, 4.0,
    4.3, 4.5, 4.0, 3.9, 4.3,
    5.2, 5.0, 5.6, 5.3, 5.1,
    5.5, 5.8, 4.9, 5.2, 5.6
  )
)
```

Take a look:

```{r}
plants
```

The data frame has 20 observations and four variables:

- `plant`
- `treatment`
- `height`
- `biomass`


# Always inspect a new dataset

Before analyzing a dataset, look at it.

These functions are good starting points.

## First rows

```{r}
head(plants)
```

## Structure

```{r}
str(plants)
```

## Dimensions

```{r}
dim(plants)
```

## Variable names

```{r}
names(plants)
```

## Summary

```{r}
summary(plants)
```

::: {.warning-box}

This is worth doing before fitting any model.

Check that R interpreted your variables the way you expected. A numeric variable accidentally imported as text, for example, can create problems later.

:::


# Accessing one variable

The `$` operator lets us access a column from a data frame.

```{r}
plants$height
```

We can then use that column in a function:

```{r}
mean(plants$height)
```

```{r}
sd(plants$height)
```

For biomass:

```{r}
mean(plants$biomass)
```


<div class="section-label">PACKAGES</div>

# Extending R with packages

Base R already does a lot, but packages add additional functions.

A few packages that come up frequently in statistics are:

| Package | Common use |
|---|---|
| `ggplot2` | Data visualization |
| `dplyr` | Data manipulation |
| `tidyr` | Reshaping data |
| `readr` | Importing data |
| `lme4` | Mixed models |
| `emmeans` | Estimated marginal means |


## Installing a package

You generally install a package once.

For example:

```{r}
#| eval: false

install.packages("ggplot2")
```

I normally run installation commands directly in the Console.


## Loading a package

Once it is installed:

```{r}
library(ggplot2)
```

You load the package again when starting a new R session.

::: {.tip-box}

A useful distinction:

**Install once**

```r
install.packages("ggplot2")
```

**Load when you use it**

```r
library(ggplot2)
```

:::


<div class="section-label section-label-pink">DATA VISUALIZATION</div>

# Getting started with ggplot2

`ggplot2` is one of the most commonly used R packages for data visualization.

At first, its syntax can look a little strange. The key is realizing that a plot is built in layers.

<div class="ggplot-formula">

<span class="formula-part">DATA</span>

<span class="formula-plus">+</span>

<span class="formula-part">AESTHETICS</span>

<span class="formula-plus">+</span>

<span class="formula-part">GEOM</span>

<span class="formula-plus">+</span>

<span class="formula-part">LABELS / THEME</span>

</div>

The basic structure looks like:

```r
ggplot(
  data,
  aes(x = ..., y = ...)
) +
  geom_...()
```

There are three main pieces.

### Data

Which dataset are we using?

```r
ggplot(plants)
```

### Aesthetics

Which variables should be represented?

```r
aes(
  x = height,
  y = biomass
)
```

### Geometry

How should those observations be drawn?

For a scatterplot:

```r
geom_point()
```

Put everything together:

```{r}
ggplot(
  plants,
  aes(
    x = height,
    y = biomass
  )
) +
  geom_point()
```


# Scatterplots

A scatterplot is useful when both variables are quantitative.

For example:

> Is plant height related to biomass?

```{r}
#| fig-width: 7
#| fig-height: 5

ggplot(
  plants,
  aes(
    x = height,
    y = biomass
  )
) +
  geom_point(
    size = 3,
    alpha = 0.8,
    color = "#13294B"
  ) +
  labs(
    title = "Plant height and biomass",
    subtitle = "Each point represents one plant",
    x = "Plant height",
    y = "Plant biomass"
  ) +
  theme_classic(
    base_size = 13
  ) +
  theme(
    plot.title = element_text(
      face = "bold",
      color = "#13294B"
    ),
    plot.subtitle = element_text(
      color = "#667085"
    )
  )
```


# Adding a regression line

We can add a fitted linear regression using `geom_smooth()`.

```{r}
#| fig-width: 7
#| fig-height: 5

ggplot(
  plants,
  aes(
    x = height,
    y = biomass
  )
) +
  geom_point(
    size = 3,
    alpha = 0.8,
    color = "#13294B"
  ) +
  geom_smooth(
    method = "lm",
    color = "#FF5F05",
    fill = "#FFB7D5",
    linewidth = 1
  ) +
  labs(
    title = "Plant height and biomass",
    subtitle = "Linear fit with 95% confidence interval",
    x = "Plant height",
    y = "Plant biomass"
  ) +
  theme_classic(
    base_size = 13
  ) +
  theme(
    plot.title = element_text(
      face = "bold",
      color = "#13294B"
    ),
    plot.subtitle = element_text(
      color = "#667085"
    )
  )
```

`method = "lm"` tells `ggplot2` to fit a linear model.

::: {.note-box}

The regression line is useful for visualizing the relationship, but the graph itself is not a substitute for statistical inference.

A formal model could be fit separately with:

```r
lm(
  biomass ~ height,
  data = plants
)
```

:::


# Histograms

A histogram is useful for looking at the distribution of one quantitative variable.

```{r}
#| fig-width: 7
#| fig-height: 5

ggplot(
  plants,
  aes(x = height)
) +
  geom_histogram(
    bins = 7,
    fill = "#FF5F05",
    color = "white",
    linewidth = 0.5
  ) +
  labs(
    title = "Distribution of plant height",
    x = "Plant height",
    y = "Number of plants"
  ) +
  theme_classic(
    base_size = 13
  ) +
  theme(
    plot.title = element_text(
      face = "bold",
      color = "#13294B"
    )
  )
```

A histogram can help us look at:

- the center of the distribution;
- spread;
- skewness;
- unusual observations;
- possible multiple modes.


# Boxplots

Suppose the question is:

> Does plant height differ between treatment groups?

We have:

- one categorical variable: `treatment`
- one quantitative variable: `height`

A boxplot is a reasonable place to start.

```{r}
#| fig-width: 7
#| fig-height: 5

ggplot(
  plants,
  aes(
    x = treatment,
    y = height
  )
) +
  geom_boxplot(
    width = 0.55,
    fill = "#FFB7D5",
    color = "#13294B",
    linewidth = 0.8
  ) +
  labs(
    title = "Plant height by treatment",
    x = NULL,
    y = "Plant height"
  ) +
  theme_classic(
    base_size = 13
  ) +
  theme(
    plot.title = element_text(
      face = "bold",
      color = "#13294B"
    )
  )
```

A boxplot shows the median, quartiles, spread, and potential outliers.


# Show the raw observations too

With a small dataset, I usually prefer seeing the individual observations.

We can combine `geom_boxplot()` and `geom_jitter()`:

```{r}
#| fig-width: 7
#| fig-height: 5

ggplot(
  plants,
  aes(
    x = treatment,
    y = height
  )
) +
  geom_boxplot(
    width = 0.52,
    fill = "#FFB7D5",
    color = "#13294B",
    linewidth = 0.8,
    outlier.shape = NA
  ) +
  geom_jitter(
    width = 0.08,
    size = 2.7,
    alpha = 0.75,
    color = "#13294B"
  ) +
  labs(
    title = "Plant height by treatment",
    subtitle = "Individual observations are shown over the boxplots",
    x = NULL,
    y = "Plant height"
  ) +
  theme_classic(
    base_size = 13
  ) +
  theme(
    plot.title = element_text(
      face = "bold",
      color = "#13294B"
    ),
    plot.subtitle = element_text(
      color = "#667085"
    )
  )
```

::: {.tip-box}

A boxplot summarizes the distribution. The points show you what was actually observed.

With small sample sizes, I find that combination much more informative than a boxplot alone.

:::


# Mapping variables to color

Color can also represent a variable.

For example:

```{r}
#| fig-width: 7
#| fig-height: 5

ggplot(
  plants,
  aes(
    x = height,
    y = biomass,
    color = treatment
  )
) +
  geom_point(
    size = 3
  ) +
  scale_color_manual(
    values = c(
      "Control" = "#13294B",
      "Fertilizer" = "#FF5F05"
    )
  ) +
  labs(
    title = "Height and biomass by treatment",
    x = "Plant height",
    y = "Plant biomass",
    color = "Treatment"
  ) +
  theme_classic(
    base_size = 13
  ) +
  theme(
    plot.title = element_text(
      face = "bold",
      color = "#13294B"
    )
  )
```


# Inside vs. outside `aes()`

This causes a lot of confusion at first.

If you write:

```r
aes(
  color = treatment
)
```

you are telling `ggplot2`:

> Use the variable `treatment` to determine color.

But if you write:

```r
geom_point(
  color = "#13294B"
)
```

you are saying:

> Make every point this same color.

So a useful rule is:

::: {.note-box}

**Inside `aes()` → a variable controls the visual property.**

**Outside `aes()` → you manually set the visual property.**

:::


# Which graph should I use?

There isn't one plot that works for every question.

A simple starting point is to identify the types of variables involved.

| Question | Variables | Useful first plot |
|---|---|---|
| What does one continuous variable look like? | Quantitative | Histogram |
| How does a measurement differ between groups? | Categorical + quantitative | Boxplot + raw points |
| Are two quantitative variables related? | Quantitative + quantitative | Scatterplot |
| How many observations are in each group? | Categorical | Bar plot |
| How does a measurement change over time? | Time + quantitative | Line plot |

The statistical question should come first. The graph comes after.


<div class="section-label">SAVING YOUR WORK</div>

# Saving a ggplot

A plot can be stored as an R object.

```{r}
height_plot <- ggplot(
  plants,
  aes(
    x = treatment,
    y = height
  )
) +
  geom_boxplot(
    width = 0.52,
    fill = "#FFB7D5",
    color = "#13294B",
    outlier.shape = NA
  ) +
  geom_jitter(
    width = 0.08,
    color = "#13294B"
  ) +
  labs(
    x = NULL,
    y = "Plant height"
  ) +
  theme_classic()

height_plot
```

Then save it with `ggsave()`:

```{r}
#| eval: false

ggsave(
  filename = "figures/plant_height.png",
  plot = height_plot,
  width = 7,
  height = 5,
  dpi = 300
)
```

I prefer saving figures through code rather than using the Export button because it makes the output reproducible.


# Scripts vs. the Console

A simple analysis script might look like:

```{r}
#| eval: false

# Load packages
library(ggplot2)

# Import data
plants <- read.csv(
  "data/plants.csv"
)

# Look at the data
head(plants)
str(plants)
summary(plants)

# Descriptive statistics
mean(
  plants$height,
  na.rm = TRUE
)

sd(
  plants$height,
  na.rm = TRUE
)

# Plot
ggplot(
  plants,
  aes(
    x = treatment,
    y = height
  )
) +
  geom_boxplot() +
  geom_jitter(
    width = 0.08
  ) +
  theme_classic()
```

Save that in something like:

```text
scripts/01_analysis.R
```


# Running code in RStudio

To run a line from a script:

### macOS

```text
Command + Enter
```

### Windows

```text
Ctrl + Enter
```

You can also highlight several lines and run them together.


# Comments

Anything after `#` is a comment.

R ignores comments when running the script.

```{r}
# Calculate the mean plant height
mean(plants$height)
```

Comments are useful for separating steps and documenting decisions.

For example:

```r
# Load required package
library(ggplot2)

# Inspect data before analysis
summary(plants)

# Compare treatment groups
ggplot(
  plants,
  aes(
    x = treatment,
    y = height
  )
) +
  geom_boxplot()
```


# Getting help

You do not need to memorize every R function.

For built-in help:

```{r}
#| eval: false

?mean
```

or:

```{r}
#| eval: false

help(mean)
```

You can do the same thing with package functions:

```{r}
#| eval: false

?geom_boxplot
```

Being able to read function documentation is much more useful than trying to memorize every argument.


<div class="section-label">WHEN SOMETHING BREAKS</div>

# A few common beginner errors

## `object not found`

You may see something like:

```text
Error: object 'Height' not found
```

Check:

- spelling;
- capitalization;
- whether the line creating the object was actually run.


## Missing quotation marks

This will not work:

```{r}
#| eval: false

name <- Maria
```

This will:

```{r}
name <- "Maria"
```


## `could not find function`

If you see:

```text
could not find function "ggplot"
```

you probably need:

```r
library(ggplot2)
```


## The Console suddenly shows `+`

Sometimes the Console changes from:

```text
>
```

to:

```text
+
```

That usually means R thinks your command is incomplete.

Common causes are a missing:

```text
)
"
}
```

Press **Esc** to cancel the command and check the syntax.


<div class="section-label section-label-pink">PRACTICE</div>

# Try it yourself

::: {.practice-box}

Suppose yield was measured for six crop varieties.

| Variety | Yield |
|---|---:|
| A | 4.2 |
| B | 5.1 |
| C | 4.8 |
| D | 5.5 |
| E | 3.9 |
| F | 5.3 |

Create the dataset:

```r
yield_data <- data.frame(
  variety = c(
    "A", "B", "C",
    "D", "E", "F"
  ),
  yield = c(
    4.2, 5.1, 4.8,
    5.5, 3.9, 5.3
  )
)
```

Then answer:

1. What is the mean yield?
2. What is the standard deviation?
3. What is the maximum yield?
4. Which variety has the highest yield?
5. Make a plot comparing the six varieties.

:::


## One possible solution

```{r}

yield_data <- data.frame(
  variety = c(
    "A", "B", "C",
    "D", "E", "F"
  ),
  yield = c(
    4.2, 5.1, 4.8,
    5.5, 3.9, 5.3
  )
)

mean(yield_data$yield)

sd(yield_data$yield)

max(yield_data$yield)

yield_data[
  which.max(yield_data$yield),
]
```

And a quick plot:

```{r}
#| fig-width: 7
#| fig-height: 5

ggplot(
  yield_data,
  aes(
    x = variety,
    y = yield
  )
) +
  geom_col(
    fill = "#FF5F05",
    width = 0.7
  ) +
  labs(
    title = "Yield by crop variety",
    x = "Variety",
    y = "Yield"
  ) +
  theme_classic(
    base_size = 13
  ) +
  theme(
    plot.title = element_text(
      face = "bold",
      color = "#13294B"
    )
  )
```


<div class="section-label">QUICK REFERENCE</div>

# A few functions worth remembering

<div class="card-grid">

<div class="info-card">

### Summarize

```r
mean(x)
median(x)
sd(x)
var(x)
range(x)
summary(x)
```

</div>

<div class="info-card">

### Inspect data

```r
head(data)
str(data)
dim(data)
names(data)
summary(data)
```

</div>

<div class="info-card">

### ggplot2

```r
ggplot()
aes()
geom_point()
geom_boxplot()
geom_histogram()
geom_jitter()
```

</div>

<div class="info-card">

### Packages

```r
install.packages()
library()
```

</div>

</div>


# Where to go next

Once these basics feel comfortable, the next useful step is learning how to import and manipulate real datasets.

That usually means getting familiar with functions such as:

```text
read_csv()
select()
filter()
mutate()
arrange()
group_by()
summarise()
```

From there, it becomes much easier to connect the R workflow to methods from a statistics class:

- t-tests;
- ANOVA;
- correlation;
- linear regression;
- multiple regression;
- mixed models.


::: {.final-box}

## A good workflow to keep in mind

**Import the data**

↓

**Inspect and clean the data**

↓

**Visualize the data**

↓

**Calculate descriptive statistics**

↓

**Choose and fit an appropriate statistical model**

↓

**Check assumptions**

↓

**Interpret the results**

The goal is not to memorize R syntax. It is to get comfortable enough with R that the code stops getting in the way of the statistics.

:::